---
title: "Markdown Demo"
output:
  pdf_document: default
  #html_document: default
---

###  Advertising data set

The `Advertising` data set contains `sales` of particular product in 200 different markets and budgets of advertisement for three different media: `TV`, `radio`, `newspaper`.

```{r}
Advertising  <- read.csv("Advertising.csv", header=TRUE)
Advertising=Advertising[,-1]
View(Advertising)
```

The data set `Advertising` has 200 rows and 4 columns. We print the first 6 rows and the last 6 rows.
```{r}
#head(Advertising)
dim(Advertising)

```




## Demo1: Interpretation of coefficients

* We wish to find relationship between `sales` and `TV`, `radio` and `newspaper`.
* We fit the model `sales ~ TV + radio + newspaper` using the `lm()` function.

```{r}

model <- lm(sales ~ TV + radio + newspaper, data=Advertising)
summary(model)
```



```{r}
names(summary(model))
```

```{r}
summary(model)$coefficients

beta_TV <- summary(model)$coefficients[2,1]
beta_TV
```



* An additional $1,000 spent on TV advertising is associated with selling approximately `r beta_TV` additional units of the product.



## Demo 2: confidence interval


We compute the 95% confidence intervals for the coefficient of TV.


```{r}
confint(model)

confint_TV = confint(model)[2,]
confint_TV
```



* An additional $1,000 spent on TV advertising is associated with selling approximately `r beta_TV` additional units of the product.
The increase is as low as  `r confint_TV[1]` units and as high as `r  confint_TV[2]` with the 95% confidence level.






